DESING AND SIMULATION SOFTWARE FOR RHODOTRON TYPE ELECTRON ACCELERATORS

### 2023-08-25 msthesis.tex
25/08/2023 update

\documentclass[a4paper,oneside,12pt]{report}
\usepackage{styles/fbe_tez}
\usepackage[utf8x]{inputenc} % To use Unicode (e.g. Turkish) characters
\usepackage{amsmath, amsthm} % Some extra symbols
\usepackage[bottom]{footmisc}
\usepackage{cite}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{multirow}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{array}
\usepackage{amssymb,bm,cite,graphicx, fixmath, texdraw}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{amsmath}
\usepackage{notoccite}
\usepackage{subcaption}
\usepackage{tabu}
\usepackage{minted}
\usepackage{xcolor}

@@

% LIST OF SYMBOLS PAGE
\begin{symbols}
    % The title will be typeset as "LIST OF SYMBOLS".
    % Use a separate \sym command for each symbols definition.
    % First, Latin symbols in alphabetical order
    \sym{$\vec{a}$}{Acceleration}
    \sym{\textbf{$\vec{B}$}}{Magnetic field}
    \sym{c}{Speed of light in vacuum}
    \sym{E}{Energy}
    \sym{\textbf{$\vec{E}$}}{Electric field}
    \sym{f}{Frequency}
    \sym{$\vec{F}$}{Force}
    \sym{$\vec{p}$}{Momentum}
    \sym{$\vec{r}$}{Position}
    \sym{t}{Time}
    \sym{T}{Period}
    \sym{$\vec{v}$}{Velocity}
    \sym{Z}{Shunt Impedance}
    % 1 EMPTY LINE BETWEEN LATIN AND GREEK SYMBOLS GROUP!!!
    \sym{}{}
    % Then Greek symbols in alphabetical order
    \sym{$\gamma$}{Lorentz Factor}
    \sym{$\beta$}{Beta Factor}
    \sym{$\phi_{lag}$}{Phase Lag}

@@

% LIST OF ABBREV PAGE
\begin{abbreviations}
    % Abbreviations in alphabetical order
   \sym{2D}{One Dimensional}
   \sym{2D}{Two Dimensional}
   \sym{3D}{Three Dimensional}
   \sym{eV}{Electron Volt}
   \sym{Hz}{Hertz}
   \sym{KAHVELab}{Kandilli Detector Accelerator and \\Instrumentation Laboratory}
   \sym{Linac}{Linear Particle Accelerator}
   \sym{ODE}{Ordinary Differential Equation}
   \sym{RF}{Radio Frequency}
   \sym{s}{Second}
   \sym{T}{Tesla}
   \sym{V}{Volt}

@@

MeV-level electron beams, both as is and as X-Ray sources, are especially useful for industrial and medical applications. 
For example, high-energy beams (5–10 MeV) are used in medical accelerators and provide services such as medical device sterilization and food irradiation;
with medium energy devices (400 keV–5 MeV) wires, cables and tires are improved, factory flue gases or wastewater can be treated.
Although most of the industrial electron accelerators operating at MeV-level are linear accelerators at the moment, due to their compact nature, circular accelerators have been gaining prominence.

Rose-pattle accelerator is a type of circular accelerator suitable for producing electron beams at MeV-level energy. 
It that was developed in 1989 and was named \textit{Rhodotron} \cite{rhodo_pottier}. This name was trademarked by \textit{IBA Group} in the following years.

@@

To start separating the parallel and perpendicular components relative to $\vecthree{\beta}$, we can find $\vecthreeBF{a}_{||}$ and $\vecthreeBF{F}_{||}$.
\begin{equation} \label{eq:a_F_parallels}
    \begin{aligned}
        \vecthreeBF{a}_{||} = \frac{(\dotProd{a}{\beta})}{\beta^2}\beta 
    \end{aligned}
    \qquad \qquad
    \begin{aligned}
        \vecthreeBF{F}_{||} = \frac{(\dotProd{F}{\beta})}{\beta^2}\beta 
    \end{aligned}
\end{equation}
\begin{eqnarray}
    \dotProd{\textbf{F}}{\beta} &=& \gamma m_0 \{ \dotProd{\textbf{a}}{\beta} + \gamma^2(\dotProd{\beta}{\textbf{a}})\beta^2 \} \\
                                &=& \gamma m_0 (\dotProd{\textbf{a}}{\beta}) \{ \gamma^2\beta^2  + 1\} 
\end{eqnarray}
Using $\gamma^2 \beta^2 + 1 = \gamma^2 $ we have,
\begin{equation}
    \dotProd{\textbf{F}}{\beta} = m_0 \gamma^3 (\dotProd{\textbf{a}}{\beta} )
\end{equation}
inserting this into \fromeq{a_F_parallels},
\begin{eqnarray}
    \vecthreeBF{F}_{||} &=& \frac{(\dotProd{\textbf{F}}{\beta})}{\beta^2} \vec{\beta} \nonumber \\
                        &=& m_0 \gamma^3 \frac{(\dotProd{\textbf{a}}{\beta} )}{\beta^2} \vec{\beta} \nonumber \\
                        &=& m_0 \gamma^3 \vecthreeBF{a}_{||} \label{eq:Frel_parallel}
\end{eqnarray}
Therefore from \fromeqs{Frel}{Frel_parallel}
\begin{eqnarray}
    \vecthreeBF{F}  &=& m_0 \gamma^3 \vecthreeBF{a}_{||} \beta^2 + m_0 \gamma \vecthreeBF{a} \\
                    &=& m_0 \gamma^3 \vecthreeBF{a}_{||} \beta^2 + m_0 \gamma \{ \vecthreeBF{a}_{||} +  \vecthreeBF{a}_{\perp} \} \\
                    &=& m_0 \vecthreeBF{a}_{||} \gamma \{ \gamma^2\beta^2 + 1 \} + m_0 \gamma \vecthreeBF{a}_{\perp} \\
                    &=& m_0 \vecthreeBF{a}_{||} \gamma^3 + m_0 \gamma \vecthreeBF{a}_{\perp} \\
                    &=& \vecthreeBF{F}_{||} + m_0 \gamma \vecthreeBF{a}_{\perp}
\end{eqnarray}
We finally have two separate equations which are in similar form with \fromeq{newton_second_law}.
\begin{equation} \label{eq:Frel_para_and_perp}
    \begin{aligned}
        \vecthreeBF{F}_{||} = \gamma^3  m_0 \vecthreeBF{a}_{||}
    \end{aligned}
    \qquad \qquad
    \begin{aligned}
        \vecthreeBF{F}_{\perp} = \gamma  m_0\vecthreeBF{a}_{\perp}
    \end{aligned}
\end{equation}

@@

\subsection{Relativistic Lorentz Force}
Similar to non-relativistic version, relativistic Lorentz Force is given by the following 4-vector equality
\begin{equation}
    \frac{\partial p^{\mu}}{\partial \tau} = q F^{\mu \nu} u_{\nu}
\end{equation}
where $ \partial \tau = \partial t / \gamma $ and greek letters are Lorentz indices,
\begin{equation}
    \begin{aligned}
        p^{\mu} = 
        \begin{bmatrix}
            W/c \\
            p_x \\
            p_y \\
            p_z
        \end{bmatrix}    
    \end{aligned}
    \qquad\qquad
    \begin{aligned}
        F^{\mu\nu} = 
        \begin{bmatrix}
                0       & -E_x/c   & -E_y/c    & -E_z/c \\
                E_x/c   &   0      & -B_z      & B_y    \\
                E_y/c   &   B_z    &  0        & -B_x   \\
                E_z/c   &   -B_y   & B_x       & 0   
        \end{bmatrix} 
    \end{aligned}
    \qquad\qquad
    \begin{aligned}
        u_{\nu} = \gamma
        \begin{bmatrix}
                 c \\
                -v_x \\
                -v_y \\
                -v_z \\
        \end{bmatrix} 
    \end{aligned}
\end{equation}
where $W$ is the energy of the particle and $\gamma$ is the Lorentz factor mentioned in the previous section.

For $ \mu = 0 $, we have the time component of the equation;
\begin{eqnarray}
    \frac{\gamma}{c}\frac{\partial W}{\partial t} &=& \frac{q \gamma \vecthree{E}\cdot \vecthree{v}}{c} = \frac{q \gamma}{c}  \frac{\vecthree{E}  \cdot \partial \vecthree{r}}{\partial t} \\
    \frac{\partial W}{\partial t} &=& q \frac{\vecthree{E}\cdot \partial \vecthree{r}} {\partial t}
\end{eqnarray}
This is the definition of work done by an electric field. For $ \mu = 1,2,3 $, we have the spacial components;
\begin{equation*}
    \frac{\partial \vecthree{p}}{\partial \tau} = \gamma \frac{\partial \vecthree{p}}{\partial t} = q \gamma (\vecthree{E} + \vecthree{v} \times \vecthree{B})
\end{equation*}
Which simplifies to non-relativistic Lorentz Force in \fromeq{lf}.

\subsection{Acceleration caused by Lorentz force}
Due to the nature of the cross product, Lorentz force caused by a magnetic field is always perpendicular to the velocity of the particle. 
Therefore the acceleration of the magnetic field is straightforward
\begin{equation*}
    \vecthreeBF{F}_B = \vecthreeBF{F}_{\perp} = \gamma  m_0\vecthreeBF{a}_{\perp} = \gamma  m_0\vecthreeBF{a}_{B}
\end{equation*}
The same thing cannot be said about electric field however. 
It can create force in any direction with respect to velocity. Therefore, we have the following equality;
\begin{equation} \label{eq:para_and_perp_acc_of_lorentz_force}
    \begin{aligned}
        \vecthreeBF{a}_{||} = \frac{q }{\gamma^3 m_0} \vecthreeBF{E}_{||}
    \end{aligned}
    \qquad \qquad
    \begin{aligned}
        \vecthreeBF{a}_{\perp} = \frac{q}{\gamma m_0} \{ \vecthreeBF{E}_{\perp} + \crossProdBF{v}{B}   \}
    \end{aligned}    
\end{equation}
Acceleration due to electric field can be simplified as;
\begin{eqnarray*}
    \vecthreeBF{a}_{\{B=0\}} = \vecthreeBF{a}_{E}   &=&  \vecthreeBF{a}_{||} + \vecthreeBF{a}_{\perp \{B=0\}} \\
                                &=& \frac{q}{m_0 \gamma} \{ \frac{\vecthreeBF{E}_{||}}{\gamma^2} + \vecthreeBF{E}_{\perp} \} \\
                                &=& \frac{q}{m_0 \gamma} \{ \{1 - \beta^2\}\vecthreeBF{E}_{||} + \vecthreeBF{E}_{\perp} \} \\
                                &=& \frac{q}{m_0 \gamma} \{ \vecthreeBF{E}_{||} + \vecthreeBF{E}_{\perp}-\beta^2\vecthreeBF{E}_{||} \} \\
                                &=& \frac{q}{m_0 \gamma} \{ \vecthreeBF{E} - \beta^2\vecthreeBF{E}_{||} \} \\
\end{eqnarray*}
Using the fact that $\vecthreeBF{E}_{||} = \vec{\beta}(\dotProd{\textbf{E}}{\beta})/\beta^2 $, we finally have
\begin{equation} \label{eq:acc_of_E_and_B}
    \begin{aligned}
        \vecthreeBF{a}_{E} = \frac{q}{ \gamma m_0} \{ \vecthreeBF{E} - \vecthreeBF{v}\frac{(\dotProdBF{E}{v})}{c^2} \}
    \end{aligned}
    \qquad \qquad
    \begin{aligned}
        \vecthreeBF{a}_{B} = \frac{q}{\gamma m_0} ( \crossProdBF{v}{B} )
    \end{aligned}
\end{equation}
where $\vecthreeBF{a}_{E}$ consists of two components in the direction of $\vecthreeBF{E}$ and $\vecthreeBF{v}$.
The component in the direction of $\vecthreeBF{v}$ depends on the $\dotProdBF{E}{v}$, 
resulting in $\vecthreeBF{a}_{E} || \vecthreeBF{E}$ if $\vecthreeBF{E} \perp \vecthreeBF{v}$.

@@

As particles gain energy in the accelerator, they approach relativistic speeds, where relativistic effects become significant. 
Special relativity governs the increase in mass and energy of the particles as they approach the speed of light, providing insights into the behavior of matter at high energies.
% TODO : ADD CIRCULAR AND LINEAR ACCELERATOR PICTURES HERE
\begin{figure}[H]
    \centering
    \includegraphics[width=.9\textwidth]{./figures/kahvelab_linac.png}
    \caption{A linear proton accelerator in KAHVELab \cite{kahvelab_linac}}
\end{figure}
\subsection{Key Concepts}

\subsubsection{Resonance Frequency}
Resonance frequency is the frequency in which the electromagnetic fields form standing waves in a cavity.
It is determined by the physical dimensions and the speed of light in the cavity's medium. 
In a rectangular cavity for example, the resonance frequency can be calculated from \fromeq{f_r_rec}.
\begin{equation}
    f_{klm} = \frac{c}{2\pi \sqrt{\mu_r \epsilon_r}} \sqrt{(\frac{k}{w})^2 + (\frac{l}{u})^2 + (\frac{m}{v})^2}
    \label{eq:f_r_rec}
\end{equation}
Where $w$, $u$, $v$ are the dimensions of the cavity, $\mu_r$ and $\epsilon_r$ are relative permability and permittivity of the cavity respectably.

@@

\subsubsection{Shunt Impedance}
The shunt impedance of an RF accelerator is a measure of the efficiency 
at which the accelerator can transform the supplied RF power into acceleration.

In addition to accelerating the particles, RF cavities are often designed to provide focusing forces. 
By carefully shaping the cavity and adjusting the electromagnetic fields, the particles can experience focusing effects as they pass through the cavity. 
This helps to maintain a tight and controlled beam. To ensure efficient acceleration, it is essential to maintain phase stability. 
This means that the particles should experience the strongest electric fields at the correct time during their passage through the cavity. 
Precise timing and synchronization of the RF power source with the particle beam are crucial to achieve phase stability and maximize energy transfer.
% TODO : ADD RF CAVITY PICTURE HERE
\begin{figure}[H]
    \centering
    \includegraphics[width=.75\textwidth]{./figures/pill_box.jpeg}
    \caption{An RF cavity used in KAHVELab}
\end{figure}

According to the Lorentz Force Law ( $\textit{Section \ref{sec:lorentz-force}}$ ), when a charged particle moves through a magnetic field, it experiences a force perpendicular to both its velocity 
vector and the magnetic field direction. This force causes the particle's trajectory to curve, resulting in a bending effect.

@@

Electrons undergo four different stages during a single pass. 
They are accelerated between the cylinderical plates and are shielded from the changing RF field while inside the inner cylinder and outside the cavity.

@@

Coaxial design of the cavity concentrates the electric field, while the magnetic field diminishes in the middle of the cylinders. 
Therefore the electrons are injected and accelerated in the plane of zero magnetic field where the electric field is strongest.
% TODO : ADD SUPERFISH AND CST SIMULATIONS OF A SIMPLE RHODOTRON CAVITY MAYBE?????

With the expectation that the electrons will accelerate to speeds $>=0.9 c$ after the first pass, a rhodotron cavity is designed so that the length of the path between successive passes is an integer multiple of $\lambda$, wavelength of the RF field ($l=p\lambda \label{eq:lpl}$).
This constraint helps with  phase stability and synchronization of the beam.

In the table below, optimized characteristics of a rhodotron cavity can be observed. 
\begin{figure}[H]
    \centering
    \includegraphics[width=.9\textwidth]{./figures/pottier_table1.png}
    \caption{Optimized characteristics of a rhodotron cavity \cite{rhodo_pottier}}
    \label{fig:pottier_table1}
\end{figure}
Here, $p$ is the integer multiplier in the equation ($l=p\lambda$) mentioned above, $R_1$ is the radius of the inner cylinder, $R_2$ is the radius of the outer cylinder, 
$Z_{se}$ is effective shunt empedance, $Z_{sp}$ is practical shunt empedance which was taken to be $0.85 Z_{se}$.

Energy gain after each pass can be calculated by the following formula \cite{rhodo_pottier}
\begin{equation}
    \label{eq:W_gain_each_pass_pottier}
    W_1 = Z_{sp}^{1/2} P^{1/2} \cos \phi  
\end{equation}
where $P$ is the dissipated power, $\phi$ is the phase lag, typically taken as $15^\circ$ \cite{rhodo_pottier}.

Considering $Z_{sp} \propto \lambda^{1/2}$, \,\,\, $W_1 \propto \lambda^{1/4}$, \,\,\,   $V \propto \lambda^3$, where $V$ is the volume of the cavity, implementing the $p=1$ design in \fromfig{pottier_table1} is much more space efficient.
\begin{figure}[H]
    \centering
    \includegraphics[width=.9\textwidth]{./figures/pottier_table2.png}
    \caption{Energy of a synchronous electron after each pass for both $p=1$ and $p=2$ \cite{rhodo_pottier}}
    \label{fig:pottier_table2}
\end{figure}
Total energy gain after $n$ passes, $W$, can then be found by \fromeq{W_gain_each_pass_pottier}, taking $\phi = 15^\circ$, $p=1$, i.e $R_2 = 0.27 \lambda$.
\begin{equation}
    \label{eq:W_total_gain_pottier}
    W \approx 2.14 \lambda^{1/4} P^{1/2} n \quad keV
\end{equation}

>>>>

\documentclass[a4paper,oneside,12pt]{report}
\usepackage[utf8x]{inputenc} % To use Unicode (e.g. Turkish) characters
\usepackage{amsmath, amsthm} % Some extra symbols
%\usepackage{subcaption}
\usepackage{styles/fbe_tez}
\usepackage[bottom]{footmisc}
\usepackage{cite}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{multirow}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{array}
\usepackage{amssymb,bm,cite,graphicx, fixmath, texdraw}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{amsmath}
\usepackage{notoccite}
\usepackage{subcaption}
\usepackage{tabu}
\usepackage{minted}
\usepackage{xcolor}

@@

% LIST OF SYMBOLS PAGE
\begin{symbols}
    % The title will be typeset as "LIST OF SYMBOLS".
    % Use a separate \sym command for each symbols definition.
    % First, Latin symbols in alphabetical order
    \sym{$\vec{a}$}{Acceleration}
    \sym{\textbf{$\vec{B}$}}{Magnetic field}
    \sym{c}{Speed of light in vacuum}
    \sym{E}{Energy}
    \sym{W}{Energy}
    \sym{\textbf{$\vec{E}$}}{Electric field}
    \sym{f}{Frequency}
    \sym{$\vec{F}$}{Force}
    \sym{$\vec{p}$}{Momentum}
    \sym{$\vec{r}$}{Position}
    \sym{t}{Time}
    \sym{T}{Period}
    \sym{$\vec{v}$}{Velocity}
    \sym{Z}{Shunt Impedance}
    % 1 EMPTY LINE BETWEEN LATIN AND GREEK SYMBOLS GROUP!!!
    \sym{}{}
    % Then Greek symbols in alphabetical order
    \sym{$\gamma$}{Lorentz Factor}
    \sym{$\beta$}{Beta Factor}
    \sym{$\phi_{lag}$}{Phase Lag}

@@

% LIST OF ABBREV PAGE
\begin{abbreviations}
    % Abbreviations in alphabetical order
   \sym{2D}{One Dimensional}
   \sym{2D}{Two Dimensional}
   \sym{3D}{Three Dimensional}
   \sym{eV}{Electron Volt}
   \sym{Hz}{Hertz}
   \sym{KAHVELab}{Kandilli Detector Accelerator and \\Instrumentation Laboratory}
   \sym{Linac}{Linear Particle Accelerator}
   \sym{ODE}{Ordinary Differential Equation}
   \sym{RF}{Radio Frequency}
   \sym{POC}{Proof of Concept}
   \sym{CPU}{Central Processing Unit}
   \sym{s}{Second}
   \sym{T}{Tesla}
   \sym{V}{Volt}
   \sym{W}{Watt}

@@

MeV-level electron beams, both as is and as X-Ray sources, are especially useful for industrial and medical applications. 
For instance,  high-energy beams (ranging from 5 to 10 MeV) find application in medical accelerators and in services such as sterilizing medical devices and irradiating food. 
Meanwhile, medium-energy devices (ranging from 400 keV to 5 MeV) contribute to the improvement of wires, cables, and tires, and also play a role in treating factory flue gases and wastewater.
Although most of the industrial electron accelerators operating at MeV-level are linear accelerators at the moment, due to their compact nature, circular accelerators have been gaining prominence.

Rose-pattle accelerator is a type of circular accelerator suitable for producing electron beams at MeV-level energy. 
It was developed in 1989 and was named \textit{Rhodotron} \cite{rhodo_pottier}. This name was trademarked by \textit{IBA Group} in the following years.

@@

To start separating the parallel and perpendicular components relative to $\vecthree{\beta}$, we can find $\vecthreeBF{a}_{||}$ and $\vecthreeBF{F}_{||}$.
\begin{equation} \label{eq:a_F_parallels}
    \begin{aligned}
        \vecthreeBF{a}_{||} = \frac{(\dotProd{a}{\beta})}{\beta^2}\beta 
    \end{aligned}
    \qquad \qquad
    \begin{aligned}
        \vecthreeBF{F}_{||} = \frac{(\dotProd{F}{\beta})}{\beta^2}\beta 
    \end{aligned}
\end{equation}
\begin{eqnarray}
    \dotProd{\textbf{F}}{\beta} &=& \gamma m_0 \{ \dotProd{\textbf{a}}{\beta} + \gamma^2(\dotProd{\beta}{\textbf{a}})\beta^2 \} \\
                                &=& \gamma m_0 (\dotProd{\textbf{a}}{\beta}) \{ \gamma^2\beta^2  + 1\} 
\end{eqnarray}
Using $\gamma^2 \beta^2 + 1 = \gamma^2 $ we have:
\begin{equation}
    \dotProd{\textbf{F}}{\beta} = m_0 \gamma^3 (\dotProd{\textbf{a}}{\beta} )
\end{equation}
inserting this into \fromeq{a_F_parallels},
\begin{eqnarray}
    \vecthreeBF{F}_{||} &=& \frac{(\dotProd{\textbf{F}}{\beta})}{\beta^2} \vec{\beta} \nonumber \\
                        &=& m_0 \gamma^3 \frac{(\dotProd{\textbf{a}}{\beta} )}{\beta^2} \vec{\beta} \nonumber \\
                        &=& m_0 \gamma^3 \vecthreeBF{a}_{||} \label{eq:Frel_parallel}
\end{eqnarray}
Therefore from \fromeqs{Frel}{Frel_parallel},
\begin{eqnarray}
    \vecthreeBF{F}  &=& m_0 \gamma^3 \vecthreeBF{a}_{||} \beta^2 + m_0 \gamma \vecthreeBF{a} \\
                    &=& m_0 \gamma^3 \vecthreeBF{a}_{||} \beta^2 + m_0 \gamma \{ \vecthreeBF{a}_{||} +  \vecthreeBF{a}_{\perp} \} \\
                    &=& m_0 \vecthreeBF{a}_{||} \gamma \{ \gamma^2\beta^2 + 1 \} + m_0 \gamma \vecthreeBF{a}_{\perp} \\
                    &=& m_0 \vecthreeBF{a}_{||} \gamma^3 + m_0 \gamma \vecthreeBF{a}_{\perp} \\
                    &=& \vecthreeBF{F}_{||} + m_0 \gamma \vecthreeBF{a}_{\perp}
\end{eqnarray}
We finally have two separate equations which are in similar form with \fromeq{newton_second_law}.
\begin{equation} \label{eq:Frel_para_and_perp}
    \begin{aligned}
        \vecthreeBF{F}_{||} = \gamma^3  m_0 \vecthreeBF{a}_{||}
    \end{aligned}
    \qquad \qquad
    \begin{aligned}
        \vecthreeBF{F}_{\perp} = \gamma  m_0\vecthreeBF{a}_{\perp}
    \end{aligned}
\end{equation}

@@

\subsection{Relativistic Lorentz Force}
Similar to non-relativistic version, relativistic Lorentz Force is given by the following 4-vector equality:
\begin{equation}
    \frac{\partial p^{\mu}}{\partial \tau} = q F^{\mu \nu} u_{\nu}
\end{equation}
where $ \partial \tau = \partial t / \gamma $ and greek letters are Lorentz indices,
\begin{equation}
    \begin{aligned}
        p^{\mu} = 
        \begin{bmatrix}
            W/c \\
            p_x \\
            p_y \\
            p_z
        \end{bmatrix}    
    \end{aligned}
    \qquad\qquad
    \begin{aligned}
        F^{\mu\nu} = 
        \begin{bmatrix}
                0       & -E_x/c   & -E_y/c    & -E_z/c \\
                E_x/c   &   0      & -B_z      & B_y    \\
                E_y/c   &   B_z    &  0        & -B_x   \\
                E_z/c   &   -B_y   & B_x       & 0   
        \end{bmatrix} 
    \end{aligned}
    \qquad\qquad
    \begin{aligned}
        u_{\nu} = \gamma
        \begin{bmatrix}
                 c \\
                -v_x \\
                -v_y \\
                -v_z \\
        \end{bmatrix} 
    \end{aligned}
\end{equation}
where $W$ is the energy of the particle and $\gamma$ is the Lorentz factor mentioned in the previous section.

For $ \mu = 0 $, we have the time component of the equation:
\begin{eqnarray}
    \frac{\gamma}{c}\frac{\partial W}{\partial t} &=& \frac{q \gamma \vecthree{E}\cdot \vecthree{v}}{c} = \frac{q \gamma}{c}  \frac{\vecthree{E}  \cdot \partial \vecthree{r}}{\partial t} \\
    \frac{\partial W}{\partial t} &=& q \frac{\vecthree{E}\cdot \partial \vecthree{r}} {\partial t}
\end{eqnarray}
This is the definition of work done by an electric field. For $ \mu = 1,2,3 $, we have the spacial components:
\begin{equation*}
    \frac{\partial \vecthree{p}}{\partial \tau} = \gamma \frac{\partial \vecthree{p}}{\partial t} = q \gamma (\vecthree{E} + \vecthree{v} \times \vecthree{B})
\end{equation*}
which simplifies to non-relativistic Lorentz Force in \fromeq{lf}.

\subsection{Acceleration caused by Lorentz force}
Due to the nature of the cross product, Lorentz force caused by a magnetic field is always perpendicular to the velocity of the particle. 
Therefore the acceleration of the magnetic field is straightforward
\begin{equation*}
    \vecthreeBF{F}_B = \vecthreeBF{F}_{\perp} = \gamma  m_0\vecthreeBF{a}_{\perp} = \gamma  m_0\vecthreeBF{a}_{B}
\end{equation*}
The same thing cannot be said about electric field however. 
It can create force in any direction with respect to velocity. Therefore, we have the following equality:
\begin{equation} \label{eq:para_and_perp_acc_of_lorentz_force}
    \begin{aligned}
        \vecthreeBF{a}_{||} = \frac{q }{\gamma^3 m_0} \vecthreeBF{E}_{||}
    \end{aligned}
    \qquad \qquad
    \begin{aligned}
        \vecthreeBF{a}_{\perp} = \frac{q}{\gamma m_0} \{ \vecthreeBF{E}_{\perp} + \crossProdBF{v}{B}   \}
    \end{aligned}    
\end{equation}
Acceleration due to electric field can be simplified as:
\begin{eqnarray*}
    \vecthreeBF{a}_{\{B=0\}} = \vecthreeBF{a}_{E}   &=&  \vecthreeBF{a}_{||} + \vecthreeBF{a}_{\perp \{B=0\}} \\
                                &=& \frac{q}{m_0 \gamma} \{ \frac{\vecthreeBF{E}_{||}}{\gamma^2} + \vecthreeBF{E}_{\perp} \} \\
                                &=& \frac{q}{m_0 \gamma} \{ \{1 - \beta^2\}\vecthreeBF{E}_{||} + \vecthreeBF{E}_{\perp} \} \\
                                &=& \frac{q}{m_0 \gamma} \{ \vecthreeBF{E}_{||} + \vecthreeBF{E}_{\perp}-\beta^2\vecthreeBF{E}_{||} \} \\
                                &=& \frac{q}{m_0 \gamma} \{ \vecthreeBF{E} - \beta^2\vecthreeBF{E}_{||} \} \\
\end{eqnarray*}
Using the fact that $\vecthreeBF{E}_{||} = \vec{\beta}(\dotProd{\textbf{E}}{\beta})/\beta^2 $, we finally have
\begin{equation} \label{eq:acc_of_E_and_B}
    \begin{aligned}
        \vecthreeBF{a}_{E} = \frac{q}{ \gamma m_0} \{ \vecthreeBF{E} - \vecthreeBF{v}\frac{(\dotProdBF{E}{v})}{c^2} \}
    \end{aligned}
    \qquad \qquad
    \begin{aligned}
        \vecthreeBF{a}_{B} = \frac{q}{\gamma m_0} ( \crossProdBF{v}{B} )
    \end{aligned}
\end{equation}
where $\vecthreeBF{a}_{E}$ consists of two components in the direction of $\vecthreeBF{E}$ and $\vecthreeBF{v}$.
The component in the direction of $\vecthreeBF{v}$ depends on the $\dotProdBF{E}{v}$, 
resulting in $\vecthreeBF{a}_{E} || \vecthreeBF{E}$ if $\vecthreeBF{E} \perp \vecthreeBF{v}$.

@@

As particles gain energy in the accelerator, they approach relativistic speeds, where relativistic effects become significant. 
Special relativity governs the increase in mass and energy of the particles as they approach the speed of light, providing insights into the behavior of matter at high energies.
% TODO : ADD CIRCULAR AND LINEAR ACCELERATOR PICTURES HERE
\begin{figure}[H]
    \centering
    \includegraphics[width=.9\textwidth]{./figures/kahvelab_linac.png}
    \caption{A linear proton accelerator in KAHVELab \cite{kahvelab_linac}}
\end{figure}

In addition to accelerating the particles, RF cavities are often designed to provide focusing forces. 
By carefully shaping the cavity and adjusting the electromagnetic fields, the particles can experience focusing effects as they pass through the cavity. 
This helps to maintain a tight and controlled beam. To ensure efficient acceleration, it is essential to maintain phase stability. 
This means that the particles should experience the strongest electric fields at the correct time during their passage through the cavity. 
Precise timing and synchronization of the RF power source with the particle beam are crucial to achieve phase stability and maximize energy transfer.
\begin{figure}[H]
    \centering
    \includegraphics[width=.75\textwidth]{./figures/pill_box.jpeg}
    \caption{An RF cavity used in KAHVELab}
\end{figure}

According to the Lorentz Force Law (\fromsec{lorentz-force}), when a charged particle moves through a magnetic field, it experiences a force perpendicular to both its velocity 
vector and the magnetic field direction. This force causes the particle's trajectory to curve, resulting in a bending effect.

\subsection{Key Concepts}

\subsubsection{Resonance Frequency}
Resonance frequency is the frequency in which the electromagnetic fields form standing waves in a cavity.
It is determined by the physical dimensions and the speed of light in the cavity's medium. 
In a rectangular cavity for example, the resonance frequency can be calculated from \fromeq{f_r_rec}.
\begin{equation}
    f_{klm} = \frac{c}{2\pi \sqrt{\mu_r \epsilon_r}} \sqrt{(\frac{k}{w})^2 + (\frac{l}{u})^2 + (\frac{m}{v})^2}
    \label{eq:f_r_rec}
\end{equation}
where $w$, $u$, $v$ are the dimensions of the cavity, $\mu_r$ and $\epsilon_r$ are relative permability and permittivity of the cavity respectably.

@@

\subsubsection{Shunt Impedance}
The shunt impedance of an RF accelerator is a measure of the efficiency 
at which the accelerator can transform the supplied RF power into acceleration.
It is defined as
\begin{equation} \label{eq:shunt}
    Z_s = \frac{V_{acc}^2}{P_{diss}}
\end{equation}
where $V_{acc}$ is the accelerating potential in which the particle is subjected to, 
$P_{diss}$ is the power dissipated on the cavity walls. 
An example \textit{shunt impedance} calculation can be found in \fromsec{cavity_of_a_rhodotron}.

@@

Electrons undergo four different stages inside the accelerator. 
They are accelerated between the cylinderical plates and are shielded from the changing RF field while inside the inner cylinder and outside the cavity.

@@

Coaxial design of the cavity concentrates the electric field, while the magnetic field diminishes in the middle of the cylinders. 
Therefore the electrons are injected and accelerated in the plane of zero magnetic field where the electric field is strongest. (\fromfig{rhodo_cavity_field_dist})

\begin{figure}[H]
    \captionsetup[subfigure]{justification=centering}
    \captionsetup{justification=centering}
    \centering
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\linewidth]{./figures/cst/cavity_E_field_dist.png}
        \caption*{$\vecthreeBF{E}$ field.}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\linewidth]{./figures/cst/cavity_B_field_dist.png}
        \caption*{$\vecthreeBF{B}$ field.}
    \end{subfigure}
    \caption{$\vecthreeBF{E}$ and $\vecthreeBF{B}$ eigenmode field distributions inside a coaxial cavity.}
    \label{fig:rhodo_cavity_field_dist}
\end{figure}
For a cavity defined by the volume between two coaxial cylinders of equal lengths ($h$) with radii of $R_1$ and $R_2$, where $R_1 < R_2$, located at the origin (\fromfig{int_curve}),
first eigenmode solution of the $E$ and $B$ fields are \cite{rhodo_pottier}
\begin{eqnarray}
    E &=& \frac{E_0}{r} \cos(\frac{\pi z}{h}) \sin(\omega t + \phi) \\
    B &=& \frac{B_0}{r} \sin(\frac{\pi z}{h}) \cos(\omega t + \phi)
\end{eqnarray}
where $\omega=2\pi f$, $f$ is the resonance frequency.

\begin{figure}[H]
    \centering
    \includegraphics[width=.6\textwidth]{./figures/illustrations/rhodo_integral_curve.pdf}
    \caption{Illustration of a simple coaxial cavity with the curve \textbf{C} in \fromeq{p_diss_C_curve}}
    \label{fig:int_curve}
\end{figure}
Because acceleration potential is located on the $z=0$ plane and $\vecthreeBF{E} \parallel \hat{r}$, $V_{acc}$ can be found by the following equation:
\begin{eqnarray} 
    V_{acc} &=& 2 \int_{R_1}^{R_2} |E|^2 dr \\
            &=& 2 E_0 \int_{R_1}^{R_2} \frac{dr}{r} \\
            &=& 2 E_0 \ln(\frac{R_2}{R_1}) \label{eq:v_acc}
\end{eqnarray}
Dissipated power $P_{diss}$, on the other hand, can be calculated as follows \cite{rf}:
\begin{eqnarray} 
    P_{diss} &=& \frac{1}{2} \int \int \rho_s | H_{||} |^2 dA \nonumber \\
             &=& \frac{\rho_s}{2\mu_0^2} \int_0^{2\pi} \int_C | B_{||} |^2 r ds d\theta \label{eq:p_diss_C_curve}
\end{eqnarray}
where $\rho_s$ is the areal skin resistivity ( $\rho_s \approx 2.51 \times 10^{-7} f^{1/2}$ for copper \cite{rhodo_pottier} ), $B=\mu H$, $\mu$ is permeability, $\mu_0$ is the vacuum permability. 
Integral curve \textbf{C} is defined as the circumference of cylindrically symmetrical cross sectional area of the cavity (\fromfig{int_curve}).
This curve can be seperated to its line components and total power dissipation of this curve, $P_C$ will be equal to sum of the power dissipated in these lines.
Since $\vecthreeBF{B}$ is always parallel to the surface we can use $\vecthreeBF{B}$ directly:
\begin{eqnarray}
    P_{diss} &=& \frac{\rho_s}{2\mu_0^2} \int_0^{2\pi} (
                   2\int_{0}^{\frac{h}{2}} |B|^2 r dz \Big|_{r=R_1}
